---
title: "r4ds: 10"
output: 
  html_notebook:
    toc: true

---
## 10 Tibbles

### 10.1 Introduction

From R for Data Science, [10: Tibbles](http://r4ds.had.co.nz/tibbles.html)
Started & Completed: 1 May, 2018

```{r}
library(tidyverse)
```

`vignette(tibble)` has more examples:

- use with list-columns

### 10.2 Creating tibbles

```{r}
as_tibble(iris)
```

```{r}
tibble(
  x = 1:5,
  y = 1,
  z = x ^ 2 + y
)
```
#### non-syntactic variable names

```{r}
tb <- tibble(
  `:)` = "smile",
  ` ` = "space",
  `2000` = "number"
)
tb
```

#### transposed tibble

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```
### 10.3 Tibbles vs. data.frame

#### printing

**Default Behaviour**

- only first 10 rows
- all columns fit on screen
- reports column type

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```


**View more**

- `print(n=[rows], width=[Inf for all])`
- `View()`: useful at end of manipulations

```{r}
nycflights13::flights %>%
  print(n=5, width = Inf)
```

**Change Tibbles Default**


- ` options(tibble.print_max = n, tibble.print_min = m) ` to show at least m, and at most n rows
- ` options(dplyr.print_min = Inf) ` to show all rows, always.
- other options listed in `package?tibble`

#### Subsetting

**Methods**

- Extract by variable name: `df$x` or `df[["x"]]`
- Extract by column position: `df[[1]]`
- in pipes: `df %>% .$x` or `df %>% .[[1]]`

** Comparison to data.frames **

- no partial matching, 
- warns if column doesn't exist
- data.frames selection `[` some times returns data.frames, or vectors
- tibbles always return tibbles

### 10.4 Interacting with older code

- some older functions don't work with `tibbles`. Use `as.data.frame()`

```{r}
class(as.data.frame(tb))
```

### 10.5 Exercises

1. tibbles have different default print behaviour

2. Compare & Contrast
    - what does `Levels:` mean?

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

```{r}
df <- tibble(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

3. You can extract a variable whose name is stored in an object using `[[obj]]`:
```{r}
var <- "abc"
df[[var]]
```

4. Practice reference to non-syntactic names
```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

# extract variable called 1
annoying$`1`

# plot scatterplot of `1` vs `2`
ggplot(data = annoying) +
  geom_point(mapping = aes( x = `1`, y = `2`))

# create new column called `3`
new_annoying <- mutate(annoying, `3` = `2` / `1`)

# rename columns to one, two, three
`colnames<-`(new_annoying,c('one','two','three'))
```

5. What does `tibble::enframe` do? When might you use it?

- `enframe()` takes named or unnamed vectors/lists and turns them into two column data frames
- `deframe()` turns two column data frams into named vector or list

6. `print(x,..., n_extra = NULL)` sets number of extra col names to print. Default is at `getOption("tibble.max_extra_cols")`

